### 2009-09-01 paper/2009-09/issues.tex
More tidying up on paper

The chained implementation of these problems
executes individual toys in the following order:
\begin{enumerate}
\item	An integer matrix $I$ with $r$ rows and $c$ columns is created
	either by using the Mandelbrot Set algorithm (Section~\ref{s:toys-mandel}) or by filling locations with random values (Section~\ref{s:toys-rng}).
	\label{e:chain-choice-1}
\item	The integer matrix $I$ is shuffled in both dimensions (Section~\ref{s:toys-half}).
\item	Either invasion percolation (Section~\ref{s:toys-invperc}) or histogram thresholding (Section~\ref{s:toys-thresh}) is used
	to generate a Boolean mask $B$ from $I$ in which the (minimum) percentage of {\tt{true}} locations is $P$.
	Like $I$, $B$ has $r$ rows and $c$ columns.
	\label{e:chain-choice-2}
\item	The Game of Life (Section~\ref{s:toys-life}) is simulated for $G$ generations, using $B$ as an initial configuration.
	This step overwrites the Boolean matrix $B$.
\item	A vector of $L$ $(m,x,y)$ points is created using the integer matrix $I$ and the Boolean matrix $B$ as described in Section~\ref{s:toys-winnow}.
\item	The elastic net algorithm is simulated using the points created in the previous step as cities.
\item	The final locations of the net points from the prevoius step are normalized, as described in Section~\ref{s:toys-norm}.
\item	An $L{\times}L$ matrix $A$ and an $L$ vector $V$ are created using the normalized point locations from the previous step,
	as described in Section~\ref{s:toys-outer}.
\item	The matrix equation $AX=V$ is solved using Gaussian elimination and using successive over-relaxation
	to generate two solution vectors $X_{gauss}$ and $X_{sor}$.
	In a parallel system, these two problems should if possible be executed concurrently.
\item	The checking vectors $V_{gauss} = A X_{gauss}$ and $V_{sor} = A X_{sor}$ are calculated.
	In a parallel system, these two problems should if possible be executed concurrently.
\item	The norm-1 distance between $V_{gauss}$ and $V_{sor}$, i.e.\ the greatest element-wise absolute difference, is calculated.
	This measures the agreement between the solutions found by the two methods.
\end{enumerate}
Because of choices in steps~\ref{e:chain-choice-1} and~\ref{e:chain-choice-2},
there are four possible chained sequences.
A minimal implementation will provide one of these,
while a full implementation will be a single program which can execute any of the four sequences.
Intermediate implementations providing all four sequences as separate programs are also acceptable.

@@

The file formats used in the Cowichan problems are shown in Figure~\ref{f:io-formats}.
A file containing a vector begins with 1 positive integer,
which specifies the number of elements in the vector;
a file containing a matrix begins with 2 positive integers,
which specify the number of rows and columns in the matrix respectively.
Elements of the vector or matrix then appear one per line,
from least-indexed to greatest-indexed
(and in row-major order for matrices).

The formats described above are required to be human-readable.
Implementations may also include I/O using ``raw'' files, i.e.\ binary data dumps rather than formatted ASCII,
using whatever file formats are convenient.
This will allow programmers to demonstrate the ``natural'' I/O capabilities of particular systems
which most probably be used for checkpointing intermediate results in real programs.

@@

\subsection{Visualization\label{s:issues-vis}}

Visualization is an increasingly important part of scientific and numerical computing.
Including it as an optional part of this problem set is also a good way to exercise both
the use of libraries within a parallel programming system,
and the ease with which a programming system can support variant versions of a program.

Our requirements for the graphics interface are:
\begin{enumerate}
\item	Visualization for individual toys should be as simple as possible,
	and should require as few changes to the basic toy code as possible.
\item	The same code should be used for both the stand-alone versions of the toys and for the chained versions.
	The latter should create all images in a single window.
\item	A single source file should implement both the graphical and non-graphical versions of each toy.
\end{enumerate}
The first two points require the system to distinguish between the actual drawing code
(e.g.\ drawing a vector of points as circles)
and the place where the drawing has to be done in the window (called the {\em{viewport}\/}).
While the stand-alone toys use a whole window for drawing,
the chained versions may only draw in some portion of the window, e.g.\ the upper left corner.
At present,
this is implemented by using a call counter
in the interfaces for the individual toys and in the chained module.
The actual drawing routines called from within each toy could then be the same.

The present visualization routines were originally written by Anil Sukul,
a student at the Vrije Universiteit, Amsterdam.
The package uses the public domain graphics library VOGLE,
available by anonymous FTP from {\tt{gondwana.ecr.mu.oz.au::/pub}}.
VOGLE includes functions for drawing points, lines, polygons, and text in a device-independent fashion, 
as well as screen control routines (such as double buffering).
VOGLE was chosen because of its simplicity,
and because it includes Pascal and Fortran interfaces.

>>>>

The chained implementation of these toys executes in the order shown below.
Because of choices in steps~\ref{e:chain-choice-1} and~\ref{e:chain-choice-2},
there are four possible chained sequences.

\begin{enumerate}

\item	An integer matrix $I$ with $r$ rows and $c$ columns is created
	either by using the Mandelbrot Set algorithm (Section~\ref{s:toys-mandel})
	or by filling locations with random values (Section~\ref{s:toys-rng}).
	\label{e:chain-choice-1}

\item	The integer matrix $I$ is shuffled in both dimensions (Section~\ref{s:toys-half}).

\item	Either invasion percolation (Section~\ref{s:toys-invperc}) or histogram thresholding (Section~\ref{s:toys-thresh}) is used
	to generate a Boolean mask $B$ from $I$ in which the (minimum) percentage of {\tt{true}} locations is $P$.
	Like $I$, $B$ has $r$ rows and $c$ columns.
	\label{e:chain-choice-2}

\item	The Game of Life (Section~\ref{s:toys-life}) is simulated for $G$ generations, using $B$ as an initial configuration.
	This step overwrites the Boolean matrix $B$.

\item	A vector of $L$ $(m,x,y)$ points is created using the integer matrix $I$ and the Boolean matrix $B$ (Section~\ref{s:toys-winnow}).

\item	A series of convex hulls are obtained from the points produced above (Section~\ref{s:toys-convex}).

\item	The final locations of the net points from the previous step are normalized (Section~\ref{s:toys-norm}).

\item	An $L{\times}L$ matrix $A$ and an $L$-vector $V$ are created using the normalized point locations from the previous step
	(Section~\ref{s:toys-outer}).

\item	The matrix equation $AX=V$ is solved using Gaussian elimination and using successive over-relaxation
	to generate two solution vectors $X_{gauss}$ and $X_{sor}$.
	These two toys should execute concurrently if possible.

\item	The checking vectors $V_{gauss}=AX_{gauss}$ and $V_{sor}=AX_{sor}$ are calculated.
	These two toys should execute concurrently if possible.

\item	The norm-1 distance between $V_{gauss}$ and $V_{sor}$ (i.e., the greatest element-wise absolute difference) is calculated.
	This measures the agreement between the solutions found by the two methods.

\end{enumerate}

@@

The file formats used in the Cowichan Problems are shown in Figure~\ref{f:io-formats}.
A file containing a vector begins with a single positive integer,
which specifies the number of elements in the vector;
a file containing a matrix begins with a pair of positive integers,
which specify the number of rows and columns in the matrix respectively\footnote{Note
that this means the first number is the Y extent, and the second number is the X extent.}.
Elements of the vector or matrix then appear one per line,
from least-indexed to greatest-indexed
(and in row-major order for matrices).

The formats described above are required to be human-readable.
Implementations may also include I/O using ``raw'' files,
i.e., binary data dumps rather than formatted ASCII,
using whatever file formats are convenient.
This will allow programmers to demonstrate the ``natural'' I/O capabilities of particular systems
which most probably be used for checkpointing intermediate results in real programs.

@@

### 2009-09-01 paper/2009-09/method.tex
More tidying up on paper

\subsection{Assessing Complexity\label{s:method-complex}}

There are at least two different ways in which usability could be compared.
One would be to measure the performance achieved by an ``average'' programmer
as a function of time on each toy.
By analogy with Hockney's $n_{1/2}$ measure \cite{b:hockney-perfparam},
we could in principle find the value of $p_{1/2}$---the programming time required to achieve
half of a machine's peak performance\footnote{Note that
	if performance was measured as a fraction of the figures quoted by manufacturers for their machines,
	it is unlikely that the halfway mark would ever be reached.}
for a particular combination of programming system and problem type.
However,
the number of implementors is likely to be too small for averages to be meaningful.

Alternatively,
we could use code metrics
to compare implementations of various toys.
This is also dubious,
as \cite{b:el-emam-metrics} has shown that code metrics do little more than measure lines of code.

We therefore intend to compare implementations through group consensus,
e.g.,
through interviews with implementors
and by asking other programmers who are familiar with the base languages used
to read and comment on the parallel versions of the toys.
We will also compare the length of code in parallel and sequential implementations,
though we realize that this can exaggerate the impact of parallelization.

>>>>

\subsection{Assessing Usability\label{s:method-assess}}

One way to compare the usability of parallel programming systems would be
to measure the performance achieved by an ``average'' programmer
as a function of time on each toy.
By analogy with Hockney's $n_{1/2}$ measure \cite{b:hockney-perfparam},
we could in principle find the value of $p_{1/2}$---the programming time required to achieve
half of a machine's peak performance\footnote{Note that
	if performance was measured as a fraction of the figures quoted by manufacturers for their machines,
	it is unlikely that the halfway mark would ever be reached.}
for a particular combination of programming system and problem type.
However,
the number of implementors required is much greater than the number we can reasonably expect.

Alternatively,
we could use code metrics
to compare implementations of various toys.
This is also dubious,
both because \cite{b:el-emam-metrics} has shown that most metrics do little more than measure lines of code,
and because of the difficulty of comparing metric values between languages.

The best we can hope for now is therefore qualitative consensus,
e.g.,
to interview implementors about their experiences
and ask other programmers who are familiar with the base languages used
to read and comment on the parallel versions of the toys.
We will also compare the length of code in parallel and sequential implementations,
though we realize that this can exaggerate the impact of parallelization.

### 2009-09-01 paper/2009-09/ops.tex
More tidying up on paper

\begin{itemize}
\item	elementwise operations on arrays (unary, binary, and scalar promotion)
\item	cumulative operations on arrays (reduction and parallel prefix)
\item	array re-shaping and re-sizing (e.g.\ sub-array extraction)
\item	partial (conditionally masked) versions of all of the above
\item	regular data motion (e.g.\ circular and planar shifting)
\item	irregular data motion (e.g.\ 1-to-1, 1-to-many, and many-to-1 permutation)
\item	scalar and non-scalar value location (e.g.\ finding a scalar or record value in an array)
\item	differential local addressing (i.e.\ subscripting an array with an array)
\item	full or partial replication of shared read-only values
\item	full or partial replication of shared read-mostly values with automatic consistency management
\item	structures with rotating ownership, suitable for migration
\item	producer-consumer structures
\item	partitionable structures
\item	pre-determined run-time re-partitioning (i.e.\ re-distributing an array)
\item	dynamic re-partitioning (e.g.\ for load balancing)
\item	committed mutual exclusion (e.g.\ waiting on a semaphore)
\item	uncommitted mutual exclusion (e.g.\ lock or fail)
\item	barrier synchronization
\item	multiple concurrent barriers used by non-overlapping groups of processes
\item	fetch-and-add, and other fetch-and-operate functions
\item	pre-scheduled receipt of messages of known composition
\item	variable-length message receipt
\item	message selection by type
\item	message selection by source
\item	message selection by contents
\item	guarded message selection (e.g.\ CSP's {\tt{alt}} construct)
\item	broadcast and partial broadcast
\item	split-phase (non-blocking) operations
\item	data marshalling and unmarshalling of ``flat'' structures (e.g.\ arrays of scalars)
\item	data marshalling and unmarshalling of nested or linked structures
\item	heterogeneous parallelism (i.e.\ running different applications concurrently in one machine)
\item	pipelining
\item	distributed linked structures in systems without physically-shared memory
\item	indexing of distributed shared arrays
\item	collective I/O
\item	uniform-size interleaved I/O
\item	heterogeneous-size interleaved I/O
\item	independent I/O operations on a single shared file
\end{itemize}

>>>>

\begin{itemize}
\item	elementwise operations on arrays (unary, binary, and scalar promotion)
\item	cumulative operations on arrays (reduction and parallel prefix)
\item	array re-shaping and re-sizing (e.g.\ sub-array extraction)
\item	partial (conditionally masked) versions of all of the above
\item	regular data motion (e.g.\ circular and planar shifting)
\item	irregular data motion (e.g.\ 1-to-1, 1-to-many, and many-to-1 permutation)
\item	scalar and non-scalar value location (e.g.\ finding a scalar or record value in an array)
\item	differential local addressing (i.e., subscripting an array with an array)
\item	full or partial replication of shared read-only values
\item	full or partial replication of shared read-mostly values with automatic consistency management
\item	structures with rotating ownership, suitable for migration
\item	producer-consumer structures
\item	partitionable structures
\item	pre-determined run-time re-partitioning (i.e., re-distributing an array)
\item	dynamic re-partitioning (e.g.\ for load balancing)
\item	committed mutual exclusion (e.g.\ waiting on a semaphore)
\item	uncommitted mutual exclusion (e.g.\ lock or fail)
\item	barrier synchronization
\item	multiple concurrent barriers used by non-overlapping groups of processes
\item	fetch-and-add, and other fetch-and-operate functions
\item	pre-scheduled receipt of messages of known composition
\item	variable-length message receipt
\item	message selection by type
\item	message selection by source
\item	message selection by contents
\item	guarded message selection (e.g.\ CSP's {\tt{alt}} construct)
\item	broadcast and partial broadcast
\item	split-phase (non-blocking) operations
\item	data marshalling and unmarshalling of ``flat'' structures (e.g.\ arrays of scalars)
\item	data marshalling and unmarshalling of nested or linked structures
\item	heterogeneous parallelism (i.e., running different applications concurrently in one machine)
\item	pipelining
\item	distributed linked structures in systems without physically-shared memory
\item	indexing of distributed shared arrays
\item	collective I/O
\item	uniform-size interleaved I/O
\item	heterogeneous-size interleaved I/O
\item	independent I/O operations on a single shared file
\end{itemize}

### 2009-09-01 paper/2009-09/prob-invperc.tex
More tidying up on paper

Invasion percolation models the displacement of one fluid
(such as oil)
by another
(such as water)
in fractured rock \cite{b:percolation-theory}.
In two dimensions,
this can be simulated by generating
an $N{\times}N$ grid of random numbers in the range $[1{\ldots}R]$,
and then marking the center cell of the grid as filled.
In each iteration,
one examines the four orthogonal neighbors of all filled cells,
chooses the one with the lowest value
(i.e.\ the one with the least resistance to filling),
and fills it in.
Figure~\ref{f:invperc1} shows the first few steps in this process,
while Figure~\ref{f:invperc2} shows the evolution of the final fractal shape.

>>>>

Invasion percolation models the displacement of one fluid
(such as oil)
by another
(such as water)
in fractured rock \cite{b:percolation-theory}.
In two dimensions,
this can be simulated by generating
an $N{\times}N$ grid of random numbers in the range $[1{\ldots}R]$,
and then marking the center cell of the grid as filled.
In each iteration,
one examines the four orthogonal neighbors of all filled cells,
chooses the one with the lowest value
(i.e., the one with the least resistance to filling),
and fills it in.
Figure~\ref{f:invperc1} shows the first few steps in this process,
while Figure~\ref{f:invperc2} shows the evolution of the final fractal shape.
